Models, derivatives, response diversity

### 2024-07-05 models_derivatives_response_diversity.Rmd
new sensitivity figure

```
##Sensitivity testing
Testing if depth-driven temperature models can adequately fit abundance data compared to a temperature and depth model.

```{r}
temp_only <- readRDS("total_models/df_pred_all.rds") |> 
  mutate(category = "temp")
temp_depth <- readRDS("total_models/depth_temp_pred_all.rds") |> 
  mutate(category = "temp_depth")

sensitivity

# OLD FIGURE 2

all_predictions <- predictions_categories |>
  ggplot(aes(temp, fit)) +
  geom_ribbon(aes(ymin = (fit - se.fit), ymax = (fit + se.fit), fill = factor(species)), alpha = 0.3) +
  geom_line(aes(color = factor(species))) +
  theme_bw() +
  facet_wrap(~species, scale = "free_y", labeller = labeller(species = label_wrap_gen(width = 10))) +
  theme(strip.background = element_rect(fill = "lightgrey"), strip.text = element_text(size = 6)) +
  scale_color_viridis(discrete = TRUE, guide = NULL, option = "G", aesthetics = c("color", "fill")) +
  xlab("Temperature") +
  ylab("Abundance")

all_predictions
# save the plot as TIFF
# tiff(paste("total_models/plots/all_predictions_models.tiff", sep = ""), units="in", width = 12, height=8, res=300)
# plot(all_predictions)
# 
# # Closing the graphical device
# dev.off()
```

```{r}
# Read in abundance and depth data of all species included for modelling
df_models <- readRDS("df_models.rds")

# find 11 most abundant  species
species_list <- df_models |> 
  filter(tot_abu > 300) |> 
  distinct(Species) |> 
  pull(Species)

# most abundant species
 # "Alburnus_alburnus", "Coregonus_sp_albeli", "Coregonus_sp","Gasterosteus_aculeatus",
 # "Gobio_gobio", "Leuciscus_leuciscus", "Perca_fluviatilis", "Rutilus_rutilus", 
 # "Sander_lucioperca", "Scardinius_erythrophthalmus","Gymnocephalus_cernua"

# filter df into the most abundant species and select the columns of interest

df_sensitivity <- df_models |> 
  filter(tot_abu > 300) |> 
  select(Lake, Species, Protocol, Depth_sample, Abundance, Presence, mean_last_7days) |> 
  #there are some NAs in depth -> drop them
  drop_na(Depth_sample, mean_last_7days)

df_sensitivity$fLake <- as.factor(df_sensitivity$Lake)
df_sensitivity$fProtocol <- as.factor(df_sensitivity$Protocol)

# Loop through the eleven species and fit temp and temp-depth model

species_list <- df_sensitivity |> 
  distinct(Species) |> 
  pull(Species)

# loop through the species, model both GAMs, check DHARMa and predictions
for (i in species_list){

  data <- df_sensitivity |> 
    filter(Species == i)

  #Model fit for temp and depth
  model1 <- gam(data = data, Abundance ~ s(mean_last_7days, k = 3) + s(Depth_sample, k = 3)
                + s(fLake, bs = 're')
                +  s(fProtocol, bs = 're'), family = ziP())
  # Model fit for temp
  model2 <- gam(data = data, Abundance ~ s(mean_last_7days, k = 3)
                + s(fLake, bs = 're')
                +  s(fProtocol, bs = 're'), family = ziP())
  # validate model predictions for temp and depth model with dharma:
  # prepare residuals
  simulationOutput <- simulateResiduals(fittedModel = model1, plot = F)
  # tiff_filename <- paste("total_models/gam_check/temp_depth_check_", i, ".tiff", sep = "")
  # tiff(tiff_filename, width = 800, height = 600)
  print(plot(simulationOutput))
  # dev.off()
  # Plotting standardized residuals against predictors
  # tiff_file_2 <- paste("total_models/gam_check/temp_depth_predictor_", i, ".tiff", sep = "")
  # tiff(tiff_file_2, width = 800, height = 600)
  print(plotResiduals(model1, data$mean_last_7days, xlab = "temp", main=NULL))
  # dev.off()

  # for predictions with random effects in the GAMs, we select a random value
  # of the level of random effects and exclude the random effects in the prediction

  # lakes where the species is present
  unique_lakes <- distinct(data, Lake) |>
    pull()

  # select random lake for prediction
  random_lake <- sample(unique_lakes, 1)

  # for temp-depth model only predict for depth where the species is present
  presence_depth <- data |> 
    filter(Presence == 1) |>
    distinct(Depth_sample) |> 
    pull(Depth_sample)

  # take randomly 5 depth values for prediction
  random_depth <- sort(sample(presence_depth, 5))

  # Convert sorted values to factors with original order
  factor_random_depth <- factor(random_depth, levels = random_depth, ordered = TRUE)

  # grid for each of the models

  # temp-depth model predicted with random depths where the species is present
 grid1 <- expand.grid(mean_last_7days = seq(
    from = min(data$mean_last_7days, na.rm = TRUE),
    to = max(data$mean_last_7days, na.rm = TRUE), length = 200),
    fProtocol = factor("VERT"), fLake = factor(random_lake), 
    Depth_sample = factor_random_depth)

  grid2 <- expand.grid(mean_last_7days = seq(
    from = min(data$mean_last_7days, na.rm = TRUE),
    to = max(data$mean_last_7days, na.rm = TRUE), length = 1000),
    fProtocol = factor("VERT"), fLake = factor(random_lake))

  # print relevant stats
  print(paste(i))
  print(glance(model1))
  print(glance(model2))

  # manually predict both models
  model_prediction1 <- predict.gam(model1, newdata = grid1,
                                  exclude = c("s(fProtocol)", "s(fLake)"), #exclude the re
                                  term = "s(mean_last_7days)", type = "response",             
                                  se.fit = TRUE)

  model_bind1 <- cbind(grid1, as.data.frame(model_prediction1)) |>
    mutate(model = factor("temp_and_depth"))

# predict model2
  model_prediction2 <- predict.gam(model2, newdata = grid2,
                                   exclude = c("s(fProtocol)", "s(fLake)"),
                                   type = "response", se.fit = TRUE)

  model_bind2 <- cbind(grid2, as.data.frame(model_prediction2)) |>
    mutate(model = factor("temp"))

  df_pred <- bind_rows(model_bind1, model_bind2) |>
    rename(temp = mean_last_7days) |>
    select(-fProtocol, -fLake) |>
    mutate(Depth_sample = factor(ifelse(is.na(Depth_sample), "temp", as.character(Depth_sample))))

  # Extract numeric depth levels
  numeric_levels <- levels(df_pred$Depth_sample)[!levels(df_pred$Depth_sample) %in% "temp"]

  # Sort numeric depth levels
  sorted_numeric_levels <- sort(as.numeric(numeric_levels))

  # Combine sorted numeric levels with "temp" level
  sorted_levels <- c(as.character(sorted_numeric_levels), "temp")

  # Reorder levels
  df_pred$Depth_sample <- factor(df_pred$Depth_sample, levels = sorted_levels)

  # Check the levels to confirm they are ordered from lowest to highest
  levels(df_pred$Depth_sample)

  # rescale predictions of both models
  df_pred_rescaled <- df_pred |>
    group_by(model) |>
    mutate(fit_rescaled = (fit - min(fit)) / (max(fit) - min(fit))) |>
    mutate(se_fit_rescaled = (se.fit - min(se.fit)) / (max(se.fit) - min(se.fit))) |> 
    mutate(species = factor(i))

  # save predictions of both models in a rds
   saveRDS(df_pred_rescaled, paste0("total_models/sensitivity/rds/sensitivity_predictions_", i, ".rds"))

# prepare total predictions of all models
sensitivity_predictions <- list.files(path = "total_models/sensitivity/rds", pattern = ".rds", full.names = TRUE) |>
  map_dfr(readRDS)

# save total predictions as RDS
saveRDS(sensitivity_predictions, "total_models/sensitivity_predictions.rds")

###Fig. S1

```{r}
# read in the sensitivity testing data
sensitivity_predictions <- readRDS("total_models/sensitivity_predictions.rds")

levels(sensitivity_predictions$species)

# Plot both model predictions together per species
 figure_s1 <- sensitivity_predictions |>
    ggplot(aes(temp, fit_rescaled, color = model, fill = model)) +
    geom_line() +
    geom_ribbon(aes(ymin = (fit_rescaled - se.fit), ymax = (fit_rescaled + se.fit)), color = NA, alpha = 0.1) +
    theme_bw() +
    # labs(title = paste("Sensitivity Test", i)) +
    scale_color_manual(values = c("#08306B", "#D73027"), aesthetics = c("color", "fill")) +
    facet_wrap(~species, scale = "free") +
    ylab("Abundance") +
    xlab("Temperature") +
   theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(), 
        legend.position = "right",
        legend.title = element_blank())

  figure_s1
  # save as TIFF
# tiff(paste("total_models/plots/sensitivity_overview.tiff", sep = ""), units="in", width=11, height=7, res=300)
# 
# plot(figure_s1)
# 
# dev.off()

```
#Derivatives calculation

>>>>

```
##Sensitivity testing
###NEW Fig. S1
Testing if depth-driven temperature models can adequately fit abundance data compared to a temperature and depth model.

```{r}
temp_only <- readRDS("total_models/df_pred_all.rds") |> 
  mutate(model = "temp")

temp_depth <- readRDS("total_models/depth_temp_pred_all.rds") |> 
  mutate(model = "temp_depth")

sensitivity_predictions <- bind_rows(temp_only, temp_depth)

supplementary_sensitivity <- sensitivity_predictions |>
    ggplot(aes(temp, fit, color = model, fill = model)) +
    geom_line() +
    # geom_ribbon(aes(ymin = (fit_rescaled - se.fit), ymax = (fit_rescaled + se.fit)), color = NA, alpha = 0.1) +
    theme_bw() +
    # labs(title = paste("Sensitivity Test", i)) +
    scale_color_manual(values = c("#D73027", "#08306B"), aesthetics = c("color", "fill")) +
     facet_wrap(~species, scale = "free", labeller = labeller(species = label_wrap_gen(width = 10))) +
    theme(strip.background = element_rect(fill = "lightgrey"), strip.text = element_text(size = 6)) +
    ylab("Abundance") +
    xlab("Temperature") +
   theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(), 
        legend.position = "right",
        legend.title = element_blank())

# # save the plot as TIFF
# tiff(paste("total_models/plots/supplementary_sensitivity_predictions.tiff", sep = ""), units="in", width = 12, height=8, res=300)
# plot(supplementary_sensitivity)
# 
# # Closing the graphical device
# dev.off()

```
####AIC comparison 
```{r}
# to be added
```

#Derivatives calculation
